An introduction to wavelets using R

### 2022-01-10 text/14/wavelets.Rmd
Worked on wavelet section of wavelets.md in 2014

@@

```{r graphing-functions}
graph_with_axis <- function(x, lab, limits=c(-1.5, 1.5)) {
  par(mar=c(2.6, 0, 0, 0))
  plot((1:2048)/2048, x, 
       xlim=c(0, 1.2), 
       ylim=limits, 
       type="l",
       axes=FALSE)
  axis(side=1, at=(0:4)/4, labels=c(1, 512, 1024, 1536, 2048))
  text(1.1, x[2048], lab, adj=0.5)
}
graph_without_axis <- function(x, lab, limits=c(-1.5, 1.5)) {
  par(mar=c(0, 0, 0, 0))
  plot((1:2048)/2048, x, 
       xlim=c(0, 1.2), 
       ylim=limits, 
       type="l",
       axes=FALSE)
  text(1.1, x[2048], lab, adj=0.5)
}
```

@@

```{r apply-fft}
ft.ecg <- fft(ecg)/2048
head(ft.ecg)
tail(ft.ecg)
ift.ecg <- fft(ft.ecg,inverse=TRUE)
head(ift.ecg)
tail(ift.ecg)
```

```{r needle-plot, eval=FALSE}
par(mar=c(0,0,0,0))
plot(1:2048, Mod(ft.ecg), axes=FALSE)
plot(Re(ft.ecg), Im(ft.ecg), axes=FALSE, type="n")
text(Re(ft.ecg), Im(ft.ecg), 1:2048)
```

The object ft.ecg contains the Fourier decomposition, and ift.ecg is the inverse of the decomposition. You need to scale the Fourier decomposition by the number of data points (2,048). The inverse gets the original signal (-0.104773, -0.081500, -0.116409,...) except that there is a "+0i" or "-0i" tacked on to the end.

@@

```{r graph-cosine-1, fig.height=1.33}
t.unit <- (1:2048)/2048
cos1 <- +Re(ft.ecg[2])*cos(2*pi*t.unit)
sin1 <- -Im(ft.ecg[2])*sin(2*pi*t.unit)
add1 <- cos1+sin1

{
  graph_without_axis(cos1, "Cosine\ncomponent", limits=range(add1))
  graph_without_axis(sin1, "Sine\ncomponent", limits=range(add1))
  graph_without_axis(add1, "Combined\ncomponent", limits=range(add1))
}
```

@@

```{r compare-to-original-curve, fig.height=1.33}
{
  graph_without_axis(add1, "Combined\ncomponent")
  graph_without_axis(ecg, "Original\nsignal")
}
graph_without_axis(add1, "Combined\ncomponent", limits=range(add1))
peak <- (3*pi/2+Arg(ft.ecg[2]))/(2*pi)
amplitude <- Mod(ft.ecg[2])
segments(peak, 0, peak, amplitude)
segments(0, 0, 1, 0)
peak
atan2(Im(ft.ecg[2]), Re(ft.ecg[2]))
atan2(Im(ft.ecg[2]), Re(ft.ecg[2]))+pi
Arg(ft.ecg[2])
(which.max(add1)-1)/2047
amplitude
max(add1)
```

@@

If you are familiar with the arctangent function, you can compute the location of peak/trough. The Mod function will tell you the amplitude.

```{r}
ft.ratio <- -Im(ft.ecg[2])/Re(ft.ecg[2])
round(2048*(atan(ft.ratio)+pi)/(2*pi))
which(add1==max(add1))
```

@@

```{r, fig.height=1.33}
ft.comp2 <- rep(0,length(ft.ecg))
ft.comp2[2] <- ft.ecg[2]
comp2 <- Re(fft(ft.comp2,inverse=TRUE))
{
  graph_without_axis(comp2, "Fourier\ncomponent", limits=range(add1))
  lines((1:2048)/2048, add1, col="red", lty="dotted")
}
```

This is the graph

ecg signal
Axis

@@

> ft.comp3 <- rep(0,length(ft.ecg))
> ft.comp3[3] <- ft.ecg[3]
> comp3 <- Re(fft(ft.comp3,inverse=TRUE))
> mx.3 <- max(comp3)
> graph.single("wave06.png",comp3,"Third\nFourier\ncomponent",1.8*mx.3*c(-1,1))
null device
1

Here is the graph

ecg signal
Axis

@@

> ft.comp4 <- rep(0,length(ft.ecg))
> ft.comp4[4] <- ft.ecg[4]
> comp4 <- Re(fft(ft.comp4,inverse=TRUE))
> mx.4 <- max(comp4)
> graph.single("wave07.png",comp4,"Fourth\nFourier\ncomponent",1.8*mx.4*c(-1,1))
null device
1

Here is the graph.

ecg signal
Axis

@@

The Fourier decomposition continues with components of higher and higher frequency. The following code computes the fiftieth component of the Fourier transform.

> ft.comp50 <- rep(0,length(ft.ecg))
> ft.comp50[50] <- ft.ecg[50]
> comp50 <- Re(fft(ft.comp50,inverse=TRUE))
> mx.50 <- max(comp50)
> mx <- 0.35*mx.50/mx.2
> graph.single("wave08.png",comp50,"Fiftieth\nFourier\ncomponent",c(-mx,mx))
null device
1

Here is the graph.

ecg signal
Axis

Here's the fiftieth Fourier component, and if you count carefully, you will get 49 peaks. Count them carefully. No, don't bother, I'll do it for you. Because that are packed so closely together, I will onlyshow you every third peak.

Counter
ecg signal
Axis

If you continue down the list, you get a higher and higher frequency or equivalently more and more peaks.
Now wavelets produce a deocomposition much like the Fourier decomposition but wavelets are different in one important way. Wavelets break the signal up into pieces and model signals within those pieces. The frequency of the wavelet signal is associated with the number of pieces. For high frequency wavelets, you break the signal up into a large number of pieces. For low frequency signals, you breah the signal up into fewer pieces.

@@

> d4.rwt <- daubcqf(4)
> d4.rwt
$h.0
[1] 0.4829629 0.8365163 0.2241439 -0.1294095

$h.1
[1] 0.1294095 0.2241439 -0.8365163 0.4829629

> wt.rwt <- mdwt(ecg,d4.rwt$h.0,L=9)
> names(wt.rwt)
[1] "y" "L"
> head(wt.rwt$y)
[,1]
[1,] 0.797545
[2,] 6.979219
[3,] -2.933776
[4,] -8.541192
[5,] 4.087582
[6,] 2.536293
> tail(wt.rwt$y)
[,1]
[2043,] 0.003011394
[2044,] -0.012342195
[2045,] 0.001102199
[2046,] -0.030707757
[2047,] -0.007824638
[2048,] 0.126029962
> wt.rwt$L
[1] 9
> iwt.rwt <- midwt(wt.rwt$y,daub4$h.0,L=9)
> names(iwt.rwt)
[1] "x" "L"
> head(iwt.rwt$x)
[,1]
[1,] -0.104773
[2,] -0.093136
[3,] -0.081500
[4,] -0.116409
[5,] -0.081500
[6,] -0.116409
> tail(iwt.rwt$x)
[,1]
[2043,] -0.430591
[2044,] -0.442227
[2045,] -0.477136
[2046,] -0.442227
[2047,] -0.477136
[2048,] -0.477136
> iwt.rwt$L
[1] 9

With the rwt package, you choose the particular wavelet decomposition using the daubcqf function. The wavelet decomposition uses the mdwt function and the inverse of the wavelet decomposition uses the midwt decomposition. The rwt is fairly minimal with a family of Duabechies wavelts, but no alternatives (such as coiflets). Don't ask me what a coiflet is.

@@

> d4.wavelets <- wt.filter("d4")
> d4.wavelets
An object of class "wt.filter"
Slot "L":
[1] 4

Slot "level":
[1] 1

Slot "h":
[1] -0.1294095 -0.2241439 0.8365163 -0.4829629

Slot "g":
[1] 0.4829629 0.8365163 0.2241439 -0.1294095

Slot "wt.class":
[1] "Daubechies"

Slot "wt.name":
[1] "d4"

Slot "transform":
[1] "dwt"

> wt.wavelets <- dwt(ecg,filter="d4")
> slotNames(wt.wavelets)
[1] "W" "V" "filter" "level" "n.boundary"
[6] "boundary" "series" "class.X" "attr.X" "aligned"
[11] "coe"
> iwt.wavelets <- idwt(wt.wavelets)
> head(iwt.wavelets)
[1] -0.10477 -0.09314 -0.08150 -0.11641 -0.08150 -0.11641
> tail(iwt.wavelets)
[1] -0.43059 -0.44223 -0.47714 -0.44223 -0.47714 -0.47714

@@

> graph.single("wave09.png",wt.comp(wt.rwt,5),"d4 wavelet")
null device
1
d4 wavelet
Axis

@@

There are four wavelets at this frequency.
> graph.single("wave10a.png",wt.comp(wt.rwt,5),"W9[1]")
null device
1
> graph.single("wave10b.png",wt.comp(wt.rwt,6),"W9[2]")
null device
1
> graph.single("wave10c.png",wt.comp(wt.rwt,7),"W9[3]")
null device
1
> graph.single("wave10d.png",wt.comp(wt.rwt,8),"W9[4]")
null device
1

Wavelet 1 of 4
Wavelet 2 of 4
Wavelet 3 of 4
Wavelet 4 of 4
Axis

@@

> graph.single("wave11a.png",wt.comp(wt.rwt, 9),"W8[1]")
null device
1
> graph.single("wave11b.png",wt.comp(wt.rwt,10),"W8[2]")
null device
1
> graph.single("wave11c.png",wt.comp(wt.rwt,11),"W8[3]")
null device
1
> graph.single("wave11d.png",wt.comp(wt.rwt,12),"W8[4]")
null device
1
> graph.single("wave11e.png",wt.comp(wt.rwt,13),"W8[5]")
null device
1
> graph.single("wave11f.png",wt.comp(wt.rwt,14),"W8[6]")
null device
1
> graph.single("wave11g.png",wt.comp(wt.rwt,15),"W8[7]")
null device
1
> graph.single("wave11h.png",wt.comp(wt.rwt,16),"W8[8]")
null device
1

Wavelet 1 of 8
Wavelet 2 of 8
Wavelet 3 of 8
Wavelet 4 of 8
Wavelet 5 of 8
Wavelet 6 of 8
Wavelet 7 of 8
Wavelet 8 of 8
Axis

@@

> graph.single("wave12a.png",wt.comp(wt.rwt,17),"W7[1]")
null device
1
> graph.single("wave12b.png",wt.comp(wt.rwt,18),"W7[2]")
null device
1
> graph.single("wave12c.png",wt.comp(wt.rwt,19),"W7[3]")
null device
1
> graph.single("wave12d.png",wt.comp(wt.rwt,20),"W7[4]")
null device
1
> graph.single("wave12e.png",wt.comp(wt.rwt,21),"W7[5]")
null device
1
> graph.single("wave12f.png",wt.comp(wt.rwt,22),"W7[6]")
null device
1
> graph.single("wave12g.png",wt.comp(wt.rwt,23),"W7[7]")
null device
1
> graph.single("wave12h.png",wt.comp(wt.rwt,24),"W7[8]")
null device
1
> graph.single("wave12i.png",wt.comp(wt.rwt,25),"W7[9]")
null device
1
> graph.single("wave12j.png",wt.comp(wt.rwt,26),"W7[10]")
null device
1
> graph.single("wave12k.png",wt.comp(wt.rwt,27),"W7[11]")
null device
1
> graph.single("wave12l.png",wt.comp(wt.rwt,28),"W7[12]")
null device
1
> graph.single("wave12m.png",wt.comp(wt.rwt,29),"W7[13]")
null device
1
> graph.single("wave12n.png",wt.comp(wt.rwt,30),"W7[14]")
null device
1
> graph.single("wave12o.png",wt.comp(wt.rwt,31),"W7[15]")
null device
1
> graph.single("wave12p.png",wt.comp(wt.rwt,32),"W7[16]")
null device
1

Wavelet 1 of 16
Wavelet 2 of 16
Wavelet 3 of 16
Wavelet 4 of 16
Wavelet 5 of 16
Wavelet 6 of 16
Wavelet 7 of 16
Wavelet 8 of 16
Wavelet 9 of 16
Wavelet 10 of 16
Wavelet 11 of 16
Wavelet 12 of 16
Wavelet 13 of 16
Wavelet 14 of 16
Wavelet 15 of 16
Wavelet 16 of 16
Axis

@@

> g <- d4.rwt$h.0
> g
[1] 0.4829629 0.8365163 0.2241439 -0.1294095

@@

> h <- d4.rwt$h.1
> h
[1] 0.1294095 0.2241439 -0.8365163 0.4829629

@@

> dot.prod(g,g)
[1] 1
> dot.prod(h,h)
[1] 1
> dot.prod(g,h)
[1] 0

@@

> dot.prod(g[1:2],g[3:4])
[1] 2.909201e-13
> dot.prod(h[1:2],h[3:4])
[1] 2.909201e-13
> dot.prod(g[1:2],h[3:4])
[1] 0
> dot.prod(h[1:2],g[3:4])
[1] 0

@@

> dot.prod(h,1:4)
[1] 1.000117e-12

@@

> wt.1 <- mdwt(ecg,g,L=1)
> V1 <- wt.1$y[ 1:1024]
> W1 <- wt.1$y[1025:2048]

@@

> comp(dot.prod(ecg, g,1),V1[1],"V1[1]")
V1[1]
By hand -0.1317145
Using R -0.1317145
> comp(dot.prod(ecg, g,3),V1[2],"V1[2]")
V1[2]
By hand -0.1399428
Using R -0.1399428
> comp(dot.prod(ecg, g,5),V1[3],"V1[3]")
V1[3]
By hand -0.148171
Using R -0.148171

@@

> comp(dot.prod(ecg,-h,1),W1[1],"W1[1]")
W1[1]
By hand 0.02247964
Using R 0.02247964
> comp(dot.prod(ecg,-h,3),W1[2],"W1[2]")
W1[2]
By hand 0.02468439
Using R 0.02468439
> comp(dot.prod(ecg,-h,5),W1[3],"W1[3]")
W1[3]
By hand -0.006023849
Using R -0.006023849

@@

A second pass applies the same pair of filters to V1. This creates 512 coefficients at the second highest frequency (W2) and 512 coefficients containing everything but W1 and W2 (V2).
> wt.2 <- mdwt(ecg,g,L=2)
> V2 <- wt.2$y[ 1: 512]
> W2 <- wt.2$y[513:1024]

Again, for convenience, separate out the first quarter (1:512) and the second quarter (513:1024) of the data.

> comp(dot.prod(V1, g,1),V2[1],"V2[1]")
V2[1]
By hand -0.1961169
Using R -0.1961169
> comp(dot.prod(V1, g,3),V2[2],"V2[2]")
V2[2]
By hand -0.1989215
Using R -0.1989215
> comp(dot.prod(V1, g,5),V2[3],"V2[3]")
V2[3]
By hand -0.1863624
Using R -0.1863624

@@

> comp(dot.prod(V1,-h,1),W2[1],"W2[1]")
W2[1]
By hand -0.00920762
Using R -0.00920762
> comp(dot.prod(V1,-h,3),W2[2],"W2[2]")
W2[2]
By hand 0.003389415
Using R 0.003389415
> comp(dot.prod(V1,-h,5),W2[3],"W2[3]")
W2[3]
By hand -0.0003374554
Using R -0.0003374554

@@

> all(wt.2$y[1025:2048]==wt.1$y[1025:2048])
[1] TRUE

@@

> wt.2a <- mdwt(V1,g,L=1)
> length(wt.2a$y)
[1] 1024
> all(wt.2a$y==wt.2$y[1:1024])
[1] TRUE

@@

> wt.3 <- mdwt(ecg,g,L=3)
> W3 <- wt.3$y[257:512]
> V3 <- wt.3$y[ 1:256]

@@

> comp(dot.prod(V2, g,1),V3[1],"V3[1]")
V3[1]
By hand -0.28061
Using R -0.28061
> comp(dot.prod(V2, g,3),V3[2],"V3[2]")
V3[2]
By hand -0.2404948
Using R -0.2404948
> comp(dot.prod(V2, g,5),V3[3],"V3[3]")
V3[3]
By hand -0.2919882
Using R -0.2919882

@@

> comp(dot.prod(V2,-h,1),W3[1],"W3[1]")
W3[1]
By hand -0.002777659
Using R -0.002777659
> comp(dot.prod(V2,-h,3),W3[2],"W3[2]")
W3[2]
By hand 0.03857375
Using R 0.03857375
> comp(dot.prod(V2,-h,5),W2[3],"W3[3]")
W3[3]
By hand 0.0963032950
Using R -0.0003374554

@@

> all(wt.3$y[ 513:1024]==wt.2$y[ 513:1024])
[1] TRUE

@@

> wt.3a <- mdwt(V2,g,L=1)
> length(wt.3a)
[1] 2
> all(wt.3a$y==wt.3$y[1:512])
[1] TRUE

@@

> W8d <- dilute(W8)
> V8d <- dilute(V8)
> comp( dot.prod(W8d,g,14)+dot.prod(V8d,h,14),V7[1],"V7[1]")
V7[1]
By hand -3.315508
Using R -3.315508
> comp(-dot.prod(W8d,g,15)-dot.prod(V8d,h,15),V7[2],"V7[2]")
V7[2]
By hand -2.398146
Using R -2.398146
> comp( dot.prod(W8d,g,16)+dot.prod(V8d,h,16),V7[3],"V7[3]")
V7[3]
By hand -0.2769286
Using R -0.2769286
> comp(-dot.prod(W8d,g, 1)-dot.prod(V8d,h, 1),V7[4],"V7[4]")
V7[4]
By hand 1.782748
Using R 1.782748
> comp( dot.prod(W8d,g, 2)+dot.prod(V8d,h, 2),V7[5],"V7[5]")
V7[5]
By hand 5.25007
Using R 5.25007
> comp(-dot.prod(W8d,g, 3)-dot.prod(V8d,h, 3),V7[6],"V7[6]")
V7[6]
By hand 4.7496
Using R 4.7496

@@

> W7d <- dilute(W7)
> V7d <- dilute(V7)
> comp( dot.prod(W7d,g,30)+dot.prod(V7d,h,30),V6[1],"V6[1]")
V6[1]
By hand -2.096562
Using R -2.096562
> comp(-dot.prod(W7d,g,31)-dot.prod(V7d,h,31),V6[2],"V6[2]")
V6[2]
By hand -2.294492
Using R -2.294492
> comp( dot.prod(W7d,g,32)+dot.prod(V7d,h,32),V6[3],"V6[3]")
V6[3]
By hand -2.441797
Using R -2.441797
> comp(-dot.prod(W7d,g, 1)-dot.prod(V7d,h, 1),V6[4],"V6[4]")
V6[4]
By hand -1.265345
Using R -1.265345
> comp( dot.prod(W7d,g, 2)+dot.prod(V7d,h, 2),V6[5],"V6[5]")
V6[5]
By hand -0.6650157
Using R -0.6650157
> comp(-dot.prod(W7d,g, 3)-dot.prod(V7d,h, 3),V6[6],"V6[6]")
V6[6]
By hand 0.08738041
Using R 0.08738041

@@

> g <- daubcqf(8)$h.0
> g
[1] 0.23037781 0.71484657 0.63088077 -0.02798377 -0.18703481 0.03084138
[7] 0.03288301 -0.01059740
> h <- -daubcqf(8)$h.1
> h
[1] -0.01059740 -0.03288301 0.03084138 0.18703481 -0.02798377 -0.63088077
[7] 0.71484657 -0.23037781

@@

> wt.img.1 <- mdwt(img,g,L=1)
> comp(sum(img[1:8,1:8]*outer(g,g)),wt.img.1$y[ 1, 1],"V1")
V1
By hand -1.580973
Using R -1.580973
> comp(sum(img[1:8,1:8]*outer(h,h)),wt.img.1$y[129,129],"W1")
W1
By hand -0.002825861
Using R -0.002825861
> comp(sum(img[1:8,1:8]*outer(h,g)),wt.img.1$y[129, 1],"W1")
W1
By hand -0.01225749
Using R -0.01225749
> comp(sum(img[1:8,1:8]*outer(g,h)),wt.img.1$y[ 1,129],"W1")
W1
By hand -0.005042734
Using R -0.005042734

>>>>

```{r setup}
knitr::opts_chunk$set(fig.height=1.33)
```

@@

```{r graphing-functions}
graph_with_axis <- function(x, lab, limits=range(x)) {
  par(mar=c(2.6, 0, 0, 0))
  plot((1:2048)/2048, x, 
       xlim=c(0, 1.2), 
       ylim=limits, 
       type="l",
       axes=FALSE)
  axis(side=1, at=(0:4)/4, labels=c(1, 512, 1024, 1536, 2048))
  text(1.1, x[2048], lab, adj=0.5)
}
graph_without_axis <- function(x, lab, limits=range(x)) {
  par(mar=c(0, 0, 0, 0))
  plot((1:2048)/2048, x, 
       xlim=c(0, 1.2), 
       ylim=limits, 
       type="l",
       axes=FALSE)
  text(1.1, x[2048], lab, adj=0.5)
}
```

@@

```{r apply-fft}
ft.ecg <- fft(ecg)/2048
head(ft.ecg)
tail(ft.ecg)
```

This function produces complex numbers, which have an interpretation in terms of sine and cosine functions.

```{r needle-plot, fig.height=5}
par(mar=c(0,0,0,0))
plot(Re(ft.ecg), Im(ft.ecg), axes=FALSE, type="n")
text(Re(ft.ecg), Im(ft.ecg), 1:2048)
```

Here is a plot of the coefficients on the complex plane. Notice that certain coefficients (1, 2, 4, 32, 2018, and 2046) are further from the origin.

### Inverse Fourier transform

```{r inverse-fft}
ift.ecg <- fft(ft.ecg,inverse=TRUE)
head(ift.ecg)
tail(ift.ecg)
```

You convert from the Fourier coefficients back to the original signal using the same fft function, but this time with an inverse=TRUE argument.

The object ft.ecg contains the Fourier decomposition, and ift.ecg is the inverse of the decomposition. You need to scale the Fourier decomposition by the number of data points (2,048). The inverse gets the original signal (`r paste0(ift.ecg[1:3], colapse=", ")`, ...) except that there is a "+0i" or "-0i" tacked on to the end.

@@

```{r graph-cosine-1}
t.unit <- (1:2048)/2048
cos1 <- +Re(ft.ecg[2])*cos(2*pi*t.unit)
sin1 <- -Im(ft.ecg[2])*sin(2*pi*t.unit)
sum1 <- cos1+sin1

{
  graph_without_axis(cos1, "Cosine\ncomponent", limits=range(sum1))
  graph_without_axis(sin1, "Sine\ncomponent", limits=range(sum1))
  graph_without_axis(sum1, "Combined\ncomponent", limits=range(sum1))
}
```

@@

```{r compare-to-original-curve}
{
  graph_without_axis(sum1, "Combined\ncomponent")
  graph_without_axis(ecg, "Original\nsignal")
}
```

@@

If you are familiar with the Mod and Arg functions, they can help you compute the phase (the location of the peak/trough) and the amplitude (the height at the peak).

@@

```{r second-component}
ft.comp2 <- ft.ecg
ft.comp2[-2] <- 0
comp2 <- Re(fft(ft.comp2,inverse=TRUE))
graph_without_axis(sum1, "Using\nInverse fft")
{
  lines((1:2048)/2048, sum1, col="red", lty="dashed")
}
```

This is the graph.

@@

The third component has sines and cosines, but the period is two.

`r Re(ft.ecg[3])` cos(2*pi*t/2)

`r Im(ft.ecg[3])` sin(2*pi*t/2).

```{r graph-third-component}
ft.comp3 <- ft.ecg
ft.comp3[-3] <- 0
comp3 <- Re(fft(ft.comp3,inverse=TRUE))
graph_without_axis(comp3, "Fourier\ncomponent")
```

@@

```{r graph-fourth-component}
ft.comp4 <- ft.ecg
ft.comp4[-4] <- 0
comp4 <- Re(fft(ft.comp4,inverse=TRUE))
graph_without_axis(comp4, "Fourier\ncomponent")
```

@@

The Fourier decomposition continues with components of higher and higher frequency. The following code computes the 32nd component of the Fourier transform.

```{r thirty-second}
ft.comp32 <- ft.ecg
ft.comp32[-32] <- 0
comp32 <- Re(fft(ft.comp32,inverse=TRUE))
graph_without_axis(comp32, "Fourier\ncomponent")
```

Here's the 32nd Fourier component, and if you count carefully, you will get 31 peaks. Count them carefully. No, don't bother, I'll do it for you. Because that are packed so closely together, I will only show you every third peak.

```{r count}
m <- max(comp32)
graph_without_axis(comp32, "Fourier\ncomponent", limits=c(-m, 1.2*m))
text((seq(1, 31, by=3)-0.5)/31, 1.2*m, seq(1, 31, by=3), col="red") 
```

If you continue down the list, you get a higher and higher frequency or equivalently more and more peaks.

### Wavelets

Now wavelets produce a deocomposition much like the Fourier decomposition but wavelets are different in one important way. Wavelets break the signal up into pieces and model signals within those pieces. The frequency of the wavelet signal is associated with the number of pieces. For high frequency wavelets, you break the signal up into a large number of pieces. For low frequency signals, you breah the signal up into fewer pieces.

@@

Perhaps the most famous of the wavelets is the Daubechies 4 wavelet.

```{r rwt-package}
d4.rwt <- daubcqf(4)
d4.rwt
wt.rwt <- mdwt(ecg,d4.rwt$h.0,L=9)
names(wt.rwt)
head(wt.rwt$y)
tail(wt.rwt$y)
wt.rwt$L
iwt.rwt <- midwt(wt.rwt$y,d4.rwt$h.0,L=9)
names(iwt.rwt)
head(iwt.rwt$x)
tail(iwt.rwt$x)
iwt.rwt$L
```

With the rwt package, you choose the particular wavelet decomposition using the daubcqf function. The wavelet decomposition uses the mdwt function and the inverse of the wavelet decomposition uses the midwt decomposition. The rwt package is fairly minimal with a family of Duabechies wavelets, but no alternatives (such as coiflets). Don't ask me what a coiflet is.

@@

```{r wavelets-package}
d4.wavelets <- wt.filter("d4")
d4.wavelets
wt.wavelets <- dwt(ecg,filter="d4")
slotNames(wt.wavelets)
iwt.wavelets <- idwt(wt.wavelets)
head(iwt.wavelets)
tail(iwt.wavelets)
```

@@

```{r graph-d4}
wt.comp <- function(x, y) {
  single_wavelet <- as.numeric((1:2048)==y)
  as.vector(midwt(single_wavelet, d4.rwt$h.0, L=9)$x)
}
graph_without_axis(wt.comp(wt.rwt,5),"d4 wavelet")
```

@@

There are four wavelets at this frequency.

```{r w9}
{
  graph_without_axis(wt.comp(wt.rwt,5),"W9[1]")
  graph_without_axis(wt.comp(wt.rwt,6),"W9[2]")
  graph_without_axis(wt.comp(wt.rwt,7),"W9[3]")
  graph_without_axis(wt.comp(wt.rwt,8),"W9[4]")
}
```

@@

```{r w8}
{
  graph_without_axis(wt.comp(wt.rwt, 9),"W8[1]")
  graph_without_axis(wt.comp(wt.rwt,10),"W8[2]")
  graph_without_axis(wt.comp(wt.rwt,11),"W8[3]")
  graph_without_axis(wt.comp(wt.rwt,12),"W8[4]")
  graph_without_axis(wt.comp(wt.rwt,13),"W8[5]")
  graph_without_axis(wt.comp(wt.rwt,14),"W8[6]")
  graph_without_axis(wt.comp(wt.rwt,15),"W8[7]")
  graph_without_axis(wt.comp(wt.rwt,16),"W8[8]")
}
```

@@

```{r s7}
{
  graph_without_axis(wt.comp(wt.rwt,17),"W7[1]")
  graph_without_axis(wt.comp(wt.rwt,18),"W7[2]")
  graph_without_axis(wt.comp(wt.rwt,19),"W7[3]")
  graph_without_axis(wt.comp(wt.rwt,20),"W7[4]")
  graph_without_axis(wt.comp(wt.rwt,21),"W7[5]")
  graph_without_axis(wt.comp(wt.rwt,22),"W7[6]")
  graph_without_axis(wt.comp(wt.rwt,23),"W7[7]")
  graph_without_axis(wt.comp(wt.rwt,24),"W7[8]")
  graph_without_axis(wt.comp(wt.rwt,25),"W7[9]")
  graph_without_axis(wt.comp(wt.rwt,26),"W7[10]")
  graph_without_axis(wt.comp(wt.rwt,27),"W7[11]")
  graph_without_axis(wt.comp(wt.rwt,28),"W7[12]")
  graph_without_axis(wt.comp(wt.rwt,29),"W7[13]")
  graph_without_axis(wt.comp(wt.rwt,30),"W7[14]")
  graph_without_axis(wt.comp(wt.rwt,31),"W7[15]")
  graph_without_axis(wt.comp(wt.rwt,32),"W7[16]")
}
```

@@

```{r high-pass}
g <- d4.rwt$h.0
g
```

@@

```{r low-pass}
h <- d4.rwt$h.1
h
```

@@

```{r orthonormal}
dot.prod <- function(x, y, offset) {
  sum(x*y)
}
dot.prod(g,g)
dot.prod(h,h)
dot.prod(g,h)
```

@@

```{r overlap}
dot.prod(g[1:2],g[3:4])
dot.prod(h[1:2],h[3:4])
dot.prod(g[1:2],h[3:4])
dot.prod(h[1:2],g[3:4])
```

@@

```{r orthogonal-to-linear}
dot.prod(h,1:4)
```

@@

```{r first-level-wavelet}
wt.1 <- mdwt(ecg,g,L=1)
V1 <- wt.1$y[ 1:1024]
W1 <- wt.1$y[1025:2048]
V1[1]
dot.prod(ecg[1:4], g)
V1[2]
dot.prod(ecg[3:6], g)
W1[1]
dot.prod(ecg[1:4], h)
W1[2]
dot.prod(ecg[3:6], h)
wt.2 <- mdwt(ecg,g,L=2)
V2 <- wt.2$y[1:512]
W2 <- wt.2$y[513:1024]
W1_check2 <- wt.2$y[1025:2048]
V2[1]
dot.prod(V1[1:4], g)
V2[2]
dot.prod(V1[3:6], g)
W2[1]
dot.prod(V1[1:4], h)
W2[2]
dot.prod(V1[3:6], h)

level_wavelet <- function(wt.rwt, i) {
  v <-wt.rwt$y
  v[(2^(i-1)):(2^i)] <- 0
  as.vector(midwt(v, d4.rwt$h.0, L=9)$x)
}
```

@@

```{r down-sampling}
dot.prod(ecg, g,1)
V1[1]
dot.prod(ecg, g,3)
V1[2]
dot.prod(ecg, g,5)
V1[3]
```

@@

```{r W1-coefficients}
dot.prod(ecg,-h,1)
W1[1]
dot.prod(ecg,-h,3)
W1[2]
dot.prod(ecg,-h,5)
W1[3]
```

@@

A second pass applies the same pair of filters to V1. This creates 512 coefficients at the second highest frequency (W2) and 512 coefficients containing everything but W1 and W2 (V2).

```{r W2-coefficients}
wt.2 <- mdwt(ecg,g,L=2)
V2 <- wt.2$y[ 1: 512]
W2 <- wt.2$y[513:1024]
dot.prod(V1, g,1)
V2[1]
dot.prod(V1, g,3)
V2[2]
dot.prod(V1, g,5)
V2[3]
```

@@

```{r V2-coefficients}
dot.prod(V1,-h,1)
W2[1]
dot.prod(V1,-h,3)
W2[2]
dot.prod(V1,-h,5)
W2[3]
```

@@

```{r}
all(wt.2$y[1025:2048]==wt.1$y[1025:2048])
```

@@

wt.2a <- mdwt(V1,g,L=1)
length(wt.2a$y)
[1] 1024
all(wt.2a$y==wt.2$y[1:1024])
[1] TRUE

@@

wt.3 <- mdwt(ecg,g,L=3)
W3 <- wt.3$y[257:512]
V3 <- wt.3$y[ 1:256]

@@

dot.prod(V2, g,1),V3[1],"V3[1]")
V3[1]
By hand -0.28061
Using R -0.28061
dot.prod(V2, g,3),V3[2],"V3[2]")
V3[2]
By hand -0.2404948
Using R -0.2404948
dot.prod(V2, g,5),V3[3],"V3[3]")
V3[3]
By hand -0.2919882
Using R -0.2919882

@@

dot.prod(V2,-h,1),W3[1],"W3[1]")
W3[1]
By hand -0.002777659
Using R -0.002777659
dot.prod(V2,-h,3),W3[2],"W3[2]")
W3[2]
By hand 0.03857375
Using R 0.03857375
dot.prod(V2,-h,5),W2[3],"W3[3]")
W3[3]
By hand 0.0963032950
Using R -0.0003374554

@@

all(wt.3$y[ 513:1024]==wt.2$y[ 513:1024])
[1] TRUE

@@

wt.3a <- mdwt(V2,g,L=1)
length(wt.3a)
[1] 2
all(wt.3a$y==wt.3$y[1:512])
[1] TRUE

@@

W8d <- dilute(W8)
V8d <- dilute(V8)
 dot.prod(W8d,g,14)+dot.prod(V8d,h,14),V7[1],"V7[1]")
V7[1]
By hand -3.315508
Using R -3.315508
-dot.prod(W8d,g,15)-dot.prod(V8d,h,15),V7[2],"V7[2]")
V7[2]
By hand -2.398146
Using R -2.398146
 dot.prod(W8d,g,16)+dot.prod(V8d,h,16),V7[3],"V7[3]")
V7[3]
By hand -0.2769286
Using R -0.2769286
-dot.prod(W8d,g, 1)-dot.prod(V8d,h, 1),V7[4],"V7[4]")
V7[4]
By hand 1.782748
Using R 1.782748
 dot.prod(W8d,g, 2)+dot.prod(V8d,h, 2),V7[5],"V7[5]")
V7[5]
By hand 5.25007
Using R 5.25007
-dot.prod(W8d,g, 3)-dot.prod(V8d,h, 3),V7[6],"V7[6]")
V7[6]
By hand 4.7496
Using R 4.7496

@@

W7d <- dilute(W7)
V7d <- dilute(V7)
 dot.prod(W7d,g,30)+dot.prod(V7d,h,30),V6[1],"V6[1]")
V6[1]
By hand -2.096562
Using R -2.096562
-dot.prod(W7d,g,31)-dot.prod(V7d,h,31),V6[2],"V6[2]")
V6[2]
By hand -2.294492
Using R -2.294492
 dot.prod(W7d,g,32)+dot.prod(V7d,h,32),V6[3],"V6[3]")
V6[3]
By hand -2.441797
Using R -2.441797
-dot.prod(W7d,g, 1)-dot.prod(V7d,h, 1),V6[4],"V6[4]")
V6[4]
By hand -1.265345
Using R -1.265345
 dot.prod(W7d,g, 2)+dot.prod(V7d,h, 2),V6[5],"V6[5]")
V6[5]
By hand -0.6650157
Using R -0.6650157
-dot.prod(W7d,g, 3)-dot.prod(V7d,h, 3),V6[6],"V6[6]")
V6[6]
By hand 0.08738041
Using R 0.08738041

@@

g <- daubcqf(8)$h.0
g
[1] 0.23037781 0.71484657 0.63088077 -0.02798377 -0.18703481 0.03084138
[7] 0.03288301 -0.01059740
h <- -daubcqf(8)$h.1
h
[1] -0.01059740 -0.03288301 0.03084138 0.18703481 -0.02798377 -0.63088077
[7] 0.71484657 -0.23037781

@@

wt.img.1 <- mdwt(img,g,L=1)
sum(img[1:8,1:8]*outer(g,g)),wt.img.1$y[ 1, 1],"V1")
V1
By hand -1.580973
Using R -1.580973
sum(img[1:8,1:8]*outer(h,h)),wt.img.1$y[129,129],"W1")
W1
By hand -0.002825861
Using R -0.002825861
sum(img[1:8,1:8]*outer(h,g)),wt.img.1$y[129, 1],"W1")
W1
By hand -0.01225749
Using R -0.01225749
sum(img[1:8,1:8]*outer(g,h)),wt.img.1$y[ 1,129],"W1")
W1
By hand -0.005042734
Using R -0.005042734
